Metabolites reflect variability introduced by mesoscale eddies in the North Pacific Subtropical Gyre

### 2024-10-24 manuscript/response_to_reviewers.txt

> Specific suggestions,
> Table 1, the format is not appropriate
Table 1 has been removed entirely.

> Figures 4, and 6 can be placed together.
While we agree that the data presented in Figures 4 and 6 can be placed side-by-side
for ease of comparison, we believe that keeping these figures in their separate
locations within the manuscript allows for better narrative flow (since the HEE
results are not discussed until several sections later) and graphical design (as
proper visualization of NMDS requires that the x- and y- axes share units).

> Reviewer 2 | 04 Oct 2024 | 12:20
> #1
> Kumler et al. report in their manuscript “Metabolites reflect variability introduced 
> by mesoscale eddies in the North Pacific Subtropical Gyre” on a very impressive 
> dataset, investigating the polar part of the meta-metabolome of a > 0.2 µm fraction 
> of the plankton. They investigate samples taken at the Meso-scope and HEE experiment 
> and discuss their data in light of the eddies found during sampling.
> It is clearly a high-quality study that opens new perspectives. Publication after 
> revision is highly recommended.
This is an accurate assessment of the manuscript.

> Sampling:
> The metabolome at different depths can be strongly affected by the sampling time. 
> Day / night variations are substantial, mainly due to diel vertical migration but 
> also to photosynthetic state of the phytoplankton. Discussing the results sampling 
> times during the transect and during sampling of the HEE should be specified and 
> statistical analysis on the impact (or lack of impact) on the data should be provided.
We have added the time of day as an explanatory factor to our analysis and reported the
new values from the PERMANOVA in the text and figures, as well as adding a supplemental
figure to visually represent some of the variance due to time of day. We do see 
diel differences at the surface and DCM during the eddy transect and these have been
reported, with the 175 meter differences more likely to be due to chance and very little
diel difference in the HEE samples at all. We have additionally added a supplemental
figure identifying a few select compounds for which diel effects are clearly visible
(trehalose and sucrose) alongside other highly abundant compounds which show no diel
effects.

> Analytics:
> The high-quality analytics is appreciated. It should be verified if the variations 
> between biological replicates could be caused by the order of measurements in the 
> sequence. If >100 samples of 15 min are run in a row there could be a drift that would 
> be hidden because of the randomization. It would be easy to have a look at the outliers 
> (e.g. Fig. 2 lower panels) are measured early / late during the sequence.
We have double-checked that our normalization to internal standards removed any
analytical drift and can confirm that while significant drift occurred over
the course of the eddy transect sample run, the normalization removed this
effect and is not a major factor in the outliers visible in Fig. 2.

### 2024-10-24 manuscript/manuscript.Rmd
Adding discussion of diel effects
New Supplemental Figure
New stats analysis (adding +time_class) to PERMANOVA
New p-values and R^2
New fig. 2 (only difference should be R2 and p-vals)

All statistics were run in R version 4.4.0. We used multivariate statistics provided by the vegan package (version 2.6-6.1, @R-vegan) to assess the overall impact of various metrics across the metabolome followed by univariate statistics to investigate individually significant responses. Non-parametric tests were used when data were unlikely to obey parametric assumptions and permutational statistics were preferred wherever possible. Multiple comparison problems were controlled where necessary using a false-discovery rate correction [@Benjamini1995]. For the analyses involving the eddy transect samples, sea level anomaly was treated as a continuous variable with the exception of the PERMANOVA, for which SLA was categorized as cyclonic if less than -5 cm and anticyclonic if greater than 5 cm, with a third category for stations in between [@Anderson2001]. Both the MESO-SCOPE eddy center and HEE analyses treated SLA as a categorical variable.

@@

The metabolome clearly differed between cyclonic and anticyclonic samples, though the magnitude of this difference varied by depth. The largest differences in particulate matter composition were detected in the DCM and 175 meter samples, as shown by the NMDS in Figure 2. At the DCM, approximately 16% of the variation in the metabolome could be explained by sea level anomaly (SLA) alone (PERMANOVA, p = 0.003). Additionally, SLA was strongly correlated with the first principal component (PC) of the metabolome  with a Pearson's *r* of 0.615 and 27% of the variance explained by PC1 (Supplemental Figure 1A), indicating that this was one of the largest sources of variation in the dataset. Notably, the samples taken from the exact center of the cyclonic eddy at Station 12 were highly distinct and likely drove much of the explained variance. In the 175 meter samples, 12% of the variance was explained by SLA (PERMANOVA, p = 0.008) and SLA was highly correlated with the second PC of the dataset (*r* = 0.697, % variance explained by PC2 = 15.4, Supplemental Figure 1A), though the first PC did not seem to have any visible pattern with metadata and appeared to largely capture variation between biological triplicates (Supplemental Figure 1B). At 15 meters, SLA trends are less evident with a larger p-value (PERMANOVA, p=0.087) and a lower R$^2$ (0.09) (Figure 2).

```{r multivar_analysis object creation - nmds/pca/permanova}
set.seed(123)
multivar_analysis <- untarg_peaks %>%
  left_join(filled_file_metadata %>% distinct(ms_run, station, filename, depth), 
          by="filename") %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), 
                      labels=c("Surface  (15 m)", "DCM  (110-120 m)", "Deep  (175 m)"))) %>%
  filter(ms_run=="MT") %>%
  nest(data=-depth) %>%
  mutate(wide_peaks=map(data, function(metab_df){
    metab_df %>%
      select(feature, filename, norm_area) %>%
      group_by(feature) %>%
      filter(sum(norm_area==0)<n()*0.9) %>%
      mutate(norm_area=scale(norm_area)[,1]) %>%
      pivot_wider(names_from = "feature", values_from = norm_area) %>%
      column_to_rownames("filename") %>%
      data.matrix()
  })) %>%
  mutate(nmds_output=map(wide_peaks, function(metab_mat){
    env_vec <- pull(filled_file_metadata, sla_corr, filename)[rownames(metab_mat)]
    vegan::metaMDS(metab_mat, k = 2, distance = "manhattan", trace = 0, 
                   autotransform = FALSE, noshare = FALSE, wascores = FALSE) %>%
    vegan::MDSrotate(env_vec)
  })) %>%
  mutate(nmds_stress=map_dbl(nmds_output, pluck, "stress")) %>%
  mutate(perm_output=map(wide_peaks, function(metab_mat){
    adon_expl <- data.frame(filename=rownames(metab_mat)) %>%
      left_join(filled_file_metadata %>% select(filename, sla_class), by="filename")
    vegan::adonis2(
      metab_mat~sla_class, data = adon_expl, by = "margin", 
      permutations = 999, method = "manhattan")
  })) %>%
  mutate(perm_r2=map_dbl(perm_output, pluck, "R2", 1)) %>%
  mutate(perm_pval=map_dbl(perm_output, pluck, "Pr(>F)", 1)) %>%
  mutate(pca_output=map(wide_peaks, prcomp)) %>%
  mutate(nmds_stress=paste("s =", round(nmds_stress, digits = 3))) %>%
  mutate(perm_r2=round(perm_r2, digits = 2)) %>%
  mutate(perm_pval=round(perm_pval, digits = 4)) %>%
  mutate(perm_label=paste0("R^2^ = ", perm_r2, "<br>p = ", perm_pval))
```

@@

```{r falkor nmds and permanova}
set.seed(123)
scaled_fk_data <- untarg_peaks %>%
  # filter(filename!="190715_Smp_FK180310S62C1-DCM_B.mzML") %>%
  left_join(filled_file_metadata %>% distinct(ms_run, station, filename, depth), 
          by="filename") %>%
  filter(ms_run=="FK") %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM"))) %>%
  select(feature, filename, norm_area) %>%
  group_by(feature) %>%
  filter(sum(norm_area==0)<n()*0.9)
metab_mat_fk <- scaled_fk_data %>%
  mutate(norm_area=scale(norm_area)[,1]) %>%
  # mutate(norm_area=rank(norm_area)) %>%
  pivot_wider(names_from = "feature", values_from = norm_area) %>%
  column_to_rownames("filename") %>%
  data.matrix()
env_vec <- pull(filled_file_metadata, sla_corr, filename)[rownames(metab_mat_fk)]
nmds_fk <- metab_mat_fk %>%
  vegan::metaMDS(k = 2, distance = "manhattan", trace = 0, autotransform = FALSE,
                 noshare = FALSE, wascores = FALSE) %>%
  vegan::MDSrotate(env_vec)
adon_expl <- data.frame(filename=rownames(metab_mat_fk)) %>%
  left_join(filled_file_metadata %>% select(filename, sla_class, depth), by="filename")
adon_fk <- vegan::adonis2(
  metab_mat_fk~sla_class+depth, data = adon_expl, by = "margin", 
  permutations = 999, method = "manhattan")
```

@@

The HEE data was characterized by high inter-replicate variability relative to the samples from the MESO-SCOPE cruise, with 25 meter samples in particular highly variable in multivariate space (Figure 6). Despite this, we still saw the importance of sea level anomaly as a significant explanatory factor in the dataset (R$^2_{SLA} = 0.090$, p-value = 0.017) as well as the larger depth differences between the surface (25 meters) and DCM (110 - 120 meters) (R$^2_{depth} = 0.251$, p-value << 0.001).

@@

When analyzed separately as distinct depths, the 25 meter samples had higher variances explained by eddy (R$^2$=0.26) and a lower p-value (0.013), while the DCM samples were no longer likely to be distinct between the two eddies (p-value = 0.171) (Figure 6, Supplemental Figure 4).

@@

In all of the datasets analyzed here, we detected a significant difference in the composition of the metabolome between the adjacent eddies. This effect was strongest in the samples taken during the Lagrangian stations at the center of each eddy in the 2017 MESO-SCOPE cruise, with nearly a quarter of the total variance explained by the eddy from which the samples were taken. In the samples taken along the eddy transect, the largest effect was detected in the deepest samples taken from 175 meters and from the DCM, with much less of a response at 25 meters. In the HEE samples, however, the opposite was true with a larger SLA response at 25 meters than that at the DCM. In each case, the differences introduced by SLA were smaller than those of depth when compared directly, with even the high-resolution sampling around the DCM finding slightly more variance explained by the 40 meter difference in sampling depth than the 40 centimeter difference in SLA. This result agrees well with prior research by @Heal2021 and @Kumler2023 showing the large effect of sampling depth on the metabolome.

@@

## Supp fig 4: HEE NMDS and PERMANOVA broken down by depth

>>>>

All statistics were run in R version 4.4.0. We used multivariate statistics provided by the vegan package (version 2.6-6.1, @R-vegan) to assess the overall impact of various metrics across the metabolome followed by univariate statistics to investigate individually significant responses. Non-parametric tests were used when data were unlikely to obey parametric assumptions and permutational statistics were preferred wherever possible. Multiple comparison problems were controlled where necessary using a false-discovery rate correction [@Benjamini1995]. For the analyses involving the eddy transect samples, sea level anomaly was treated as a continuous variable with the exception of the PERMANOVA, for which SLA was categorized as cyclonic if less than -5 cm and anticyclonic if greater than 5 cm, with a third category for stations in between [@Anderson2001]. Both the MESO-SCOPE eddy center and HEE analyses treated SLA as a categorical variable. Where time of day was included as an explanatory factor in the PERMANOVA, it was binned into 6-hour increments starting from 3AM (e.g. Midday contains casts collected between 9AM and 3PM).

@@

The metabolome clearly differed between cyclonic and anticyclonic samples, though the magnitude of this difference varied by depth. The largest differences in particulate matter composition were detected in the DCM and 175 meter samples, as shown by the NMDS in Figure 2.

At the DCM, approximately 14% of the variation in the metabolome could be explained by sea level anomaly (SLA) alone (PERMANOVA, p = 0.002). Additionally, SLA was strongly correlated with the first principal component (PC) of the metabolome  with a Pearson's *r* of 0.615 and 27% of the variance explained by PC1 (Supplemental Figure 1A), indicating that this was one of the largest sources of variation in the dataset. Notably, the samples taken from the exact center of the cyclonic eddy at Station 12 were highly distinct and likely drove much of the explained variance.

In the 175 meter samples, 13% of the variance was explained by SLA (PERMANOVA, p = 0.004) and SLA was highly correlated with the second PC of the dataset (*r* = 0.697, % variance explained by PC2 = 15.4, Supplemental Figure 1A), though the first PC did not seem to have any visible pattern with metadata and appeared to largely capture variation between biological triplicates (Supplemental Figure 1B).

At 15 meters, SLA trends are slightly less evident with a larger p-value (PERMANOVA, p=0.010) and a lower R$^2$ (0.12) (Figure 2). Time of day was also a significant factor in the PERMANOVA at the surface and DCM (p-values of 0.004 and 0.032, respectively) while the diel effects in the 175 meter samples were unsurprisingly much lower (p = 0.118).

The significance of SLA as an explanatory factor was similar even when excluding time of day, with the surface seeing the largest change to an R$^2_{SLA}$ of 0.09 and a p-value of 0.087, though the fraction of the variance explained by SLA increased to 16% without time of day as a factor.

```{r multivar_analysis object creation - nmds/pca/permanova}
set.seed(123)
multivar_analysis <- untarg_peaks %>%
  left_join(filled_file_metadata %>% distinct(ms_run, station, filename, depth), 
          by="filename") %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"), 
                      labels=c("Surface  (15 m)", "DCM  (110-120 m)", "Deep  (175 m)"))) %>%
  filter(ms_run=="MT") %>%
  arrange(depth) %>%
  nest(data=-depth) %>%
  mutate(wide_peaks=map(data, function(metab_df){
    metab_df %>%
      select(feature, filename, norm_area) %>%
      group_by(feature) %>%
      filter(sum(norm_area==0)<n()*0.9) %>%
      mutate(norm_area=scale(norm_area)[,1]) %>%
      pivot_wider(names_from = "feature", values_from = norm_area) %>%
      column_to_rownames("filename") %>%
      data.matrix()
  })) %>%
  mutate(nmds_output=map(wide_peaks, function(metab_mat){
    env_vec <- pull(filled_file_metadata, sla_corr, filename)[rownames(metab_mat)]
    vegan::metaMDS(metab_mat, k = 2, distance = "manhattan", trace = 0, 
                   autotransform = FALSE, noshare = FALSE, wascores = FALSE) %>%
    vegan::MDSrotate(env_vec)
  })) %>%
  mutate(nmds_stress=map_dbl(nmds_output, pluck, "stress")) %>%
  mutate(perm_output=map(wide_peaks, function(metab_mat){
    adon_expl <- data.frame(filename=rownames(metab_mat)) %>%
      left_join(filled_file_metadata %>% select(filename, sla_class, time_class),
                by="filename")
    vegan::adonis2(
      metab_mat~sla_class, 
      # metab_mat~sla_class+time_class, 
      data = adon_expl, by = "margin", 
      permutations = 999, method = "manhattan")
  })) %>%
  mutate(perm_r2=map_dbl(perm_output, pluck, "R2", 1)) %>%
  mutate(perm_pval=map_dbl(perm_output, pluck, "Pr(>F)", 1)) %>%
  mutate(pca_output=map(wide_peaks, prcomp)) %>%
  mutate(nmds_stress=paste("s =", round(nmds_stress, digits = 3))) %>%
  mutate(perm_r2=round(perm_r2, digits = 2)) %>%
  mutate(perm_pval=round(perm_pval, digits = 4)) %>%
  mutate(perm_label=paste0("R^2^ = ", perm_r2, "<br>p = ", perm_pval))
```

@@

```{r falkor nmds and permanova}
set.seed(123)
scaled_fk_data <- untarg_peaks %>%
  # filter(filename!="190715_Smp_FK180310S62C1-DCM_B.mzML") %>%
  left_join(filled_file_metadata %>% distinct(ms_run, station, filename, depth), 
          by="filename") %>%
  filter(ms_run=="FK") %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM"))) %>%
  select(feature, filename, norm_area) %>%
  group_by(feature) %>%
  filter(sum(norm_area==0)<n()*0.9)
metab_mat_fk <- scaled_fk_data %>%
  mutate(norm_area=scale(norm_area)[,1]) %>%
  # mutate(norm_area=rank(norm_area)) %>%
  pivot_wider(names_from = "feature", values_from = norm_area) %>%
  column_to_rownames("filename") %>%
  data.matrix()
env_vec <- pull(filled_file_metadata, sla_corr, filename)[rownames(metab_mat_fk)]
nmds_fk <- metab_mat_fk %>%
  vegan::metaMDS(k = 2, distance = "manhattan", trace = 0, autotransform = FALSE,
                 noshare = FALSE, wascores = FALSE) %>%
  vegan::MDSrotate(env_vec)
adon_expl <- data.frame(filename=rownames(metab_mat_fk)) %>%
  left_join(filled_file_metadata %>% select(filename, sla_class, depth), by="filename")
adon_fk <- vegan::adonis2(
  metab_mat_fk~sla_class+depth, data = adon_expl, by = "margin", 
  permutations = 999, method = "manhattan")
# vegan::adonis2(
#   metab_mat_fk~sla_class+depth+time_class, data = adon_expl, by = "margin",
#   permutations = 999, method = "manhattan")
```

@@

The HEE data was characterized by high inter-replicate variability relative to the samples from the MESO-SCOPE cruise, with 25 meter samples in particular highly variable in multivariate space (Figure 6). Despite this, we still saw the importance of sea level anomaly as a significant explanatory factor in the dataset (R$^2_{SLA} = 0.090$, p-value = 0.017) as well as the larger depth differences between the surface (25 meters) and DCM (110 - 120 meters) (R$^2_{depth} = 0.251$, p-value << 0.001). We saw no difference between the 6AM and 6PM samples (R$^2_{time} = 0.031$, p-value = 0.373) despite a few known compounds (trehalose and sucrose) showing large differences, likely due to a majority of the mass features demonstrating no diel effect (Supplemental Figure 4).

@@

When analyzed separately as distinct depths, the 25 meter samples had higher variances explained by eddy (R$^2$=0.26) and a lower p-value (0.013), while the DCM samples were no longer likely to be distinct between the two eddies (p-value = 0.171) (Figure 6, Supplemental Figure 5).

@@

In all of the datasets analyzed here, we detected a significant difference in the composition of the metabolome between the adjacent eddies. This effect was strongest in the samples taken during the Lagrangian stations at the center of each eddy in the 2017 MESO-SCOPE cruise, with nearly a quarter of the total variance explained by the eddy from which the samples were taken. In the samples taken along the eddy transect, the largest effect was detected in the deepest samples taken from 175 meters and from the DCM, with much less of a response at 25 meters. In the HEE samples, however, the opposite was true with a surprisingly larger SLA response at 25 meters than that at the DCM. In each case, the differences introduced by SLA were smaller than those of depth when compared directly, with even the high-resolution sampling around the DCM finding slightly more variance explained by the 40 meter difference in sampling depth than the 40 centimeter difference in SLA. This result agrees well with prior research by @Heal2021 and @Kumler2023 showing the large effect of sampling depth on the metabolome. Time of day was included as a significant factor in the MESO-SCOPE transect, with trehalose and sucrose in particular showing large differences between 6AM and 6PM as has been noted previously [@Muratore2022], though this effect was diluted at 175 meters.

@@

## Supp fig 4: Boxplots of selected compounds by depth, time, and SLA

```{r diel, include=FALSE}
filled_file_metadata %>% 
  filter(samp_type=="Smp") %>%
  filter(ms_run=="MT") %>%
  distinct(station, time_class) %>%
  count(time_class)
filled_file_metadata %>% 
  filter(samp_type=="Smp") %>%
  filter(ms_run=="MT") %>%
  distinct(station, sla_class) %>%
  count(sla_class)

select_sla_time_boxplot_gp <- targ_peaks %>%
  filter(compound_name%in%c("Sucrose", "Trehalose", "Guanine", "Glycine betaine", "L-Glutamine", "Ectoine")) %>%
  mutate(compound_name=case_when(
    compound_name=="Glycine betaine"~"Glycine\nbetaine",
    compound_name=="L-Glutamine"~"Glutamine",
    TRUE~compound_name
  )) %>%
  select(filename, compound_name, nM) %>%
  left_join(filled_file_metadata) %>%
  filter(samp_type=="Smp") %>%
  filter(ms_run!="MC") %>%
  select(filename, compound_name, cruise, depth, sla_class, time_class, nM) %>%
  mutate(time_class=factor(time_class, levels=c("Morning", "Midday", "Evening", "Night"))) %>%
  mutate(cruise=factor(cruise, levels=c("MS", "FK"), labels=c("MESO-SCOPE transect", "HEE"))) %>%
  mutate(depth=factor(depth, levels=c("Surface", "DCM", "Deep"))) %>%
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_boxplot(aes(x=time_class, y=nM, color=sla_class)) +
  facet_nested(compound_name~cruise+depth, scales="free", space = "free_x") +
  labs(x=NULL, y="Metabolite concentration (nM)", color="SLA condition") +
  scale_color_manual(breaks = c("Cyclone", "Neither", "Anticyclone"),
                     values = c(scales::muted("blue"), "grey50", 
                                scales::muted("red"))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5))

ggsave("diel_effect_boxplots.png", plot = select_sla_time_boxplot_gp, device = "png",
       path = "supplement", width = 6.875, height = 6, units = "in", dpi = 300,
       bg="white")
```

![](supplement/diel_effect_boxplots.png)

*Supplemental Figure 4: Boxplots showing the concentrations of six selected compounds between cruises and depths as well as time and SLA classes. The abscissa is separated by the time category into which sampling was binned, referring to 6-hour increments starting from 3AM (e.g. Midday contains casts collected between 9AM and 3PM). Colors denote the sea level anomaly bins with a ±5cm cutoff. For the eddy transect, six stations are included in the anticyclone, three in the cyclone, and two were neither. There were two morning stations, 4 midday stations, 2 evening stations, and 3 night stations. Diel effects are clearly visible for sucrose and trehalose while other compounds show very little diel differences. Trehalose was not quantified during the MESO-SCOPE transect for lack of an authentic standard to quantify the response factor.*

## Supp fig 5: HEE NMDS and PERMANOVA broken down by depth
